---
title: 'Advent Of Code: 2017-03'
author: "Tan Ho"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(glue)
  
  knitr::opts_chunk$set(echo = TRUE)
})

options(scipen = 9999999)
options(dplyr.summarise.inform = FALSE)
```

--- Data ---

```{r eval = FALSE}
# tanho63/aoc.elf
aoc.elf::aoc_get(day = 3, year = 2017)
```

```{r}
input <- readLines(here::here("2017/day-03-input.txt")) |> as.numeric()
```

--- Part 1 ---

Recognize bottom right hand corner numbers increase by a square of odd numbers sequence. Find smallest square less than input, then start subtracting the square's size to figure out which edge of the box the input is on. Mine was on the top edge because the difference was > than square_size. Determine xloc, yloc is 1, and because we add a new top row, shift the midpoint from the 295,295 square down one y location. 

```{r}
square_size <- floor(sqrt(input))
difference <- input - square_size^2
x_loc <- (square_size+1) - (difference - (square_size+1))
y_loc <- 1

middle <- c(295,296)
sum(middle-c(x_loc,y_loc))
```

--- Part 2 ---

```{r}

size <- 5

generate_ring <- function(size){
  stopifnot(size%%2 == 1)
  
  a <- (size-2)^2+1
  b <- size^2
  b:a
}

```

