---
title: "Adv2020 - Day Ten"
author: "Tan Ho"
date: "2020-12-10"
output: html_document
---

Day ten - Adapter Array!

```{r setup, message = FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE)
```

> Find a chain that uses all of your adapters to connect the charging outlet to your device's built-in adapter and count the joltage differences between the charging outlet, the adapters, and your device. What is the number of 1-jolt differences multiplied by the number of 3-jolt differences?


--- Data ---
```{r}
input_10 <- read_lines(here("2020","day-10.txt"))
```

--- Cleaning ---
```{r}
df_adapters <- tibble(adapter = input_10) %>% 
  mutate(adapter = as.numeric(adapter)) %>% 
  add_row(adapter = max(.$adapter)+3) %>% 
  arrange(adapter) %>% 
  mutate(diff = adapter - lag(adapter,default = 0))
```

--- Problem 1 ---

```{r}

p1 <- df_adapters %>% 
  count(diff)

prod(p1$n)
```

--- Problem 2 ---

What is the total number of distinct ways you can arrange the adapters to connect the charging outlet to your device?

```{r}

library(slider)
options(scipen = 999)

p2 <- df_adapters %>%
  select(adapter) %>%
  add_row(adapter = 0) %>% 
  arrange(desc(adapter)) %>% 
  mutate(
    ways = slide(adapter,~.x,.before = 3,.after = -1),
    ways = map2(ways,adapter,~.x[.x-.y<=3])) %>% 
  tail(-1)

path_keys <- numeric()
path_keys[[max(df_adapters$adapter)]] <- 1

sum_ways <- function(ways,adapter){
  
  ways <- unlist(ways)
  
  paths <- sum(path_keys[ways])
  path_keys[adapter] <<- paths
  paths
}

p2_solve <- p2 %>% 
  mutate(paths = map2_dbl(ways,adapter,sum_ways))

max(p2_solve$paths)


```

