---
title: "Adv2020 - Day Eleven"
author: "Tan Ho"
date: "2020-12-11"
output: html_document
---

Day eleven - seating system!

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(furrr)
  library(memoise)
  
  plan(multisession)
  
  knitr::opts_chunk$set(echo = TRUE)
})
```

--- Description ---

> All decisions are based on the number of occupied seats adjacent to a given seat (one of the eight positions immediately up, down, left, right, or diagonal from the seat). The following rules are applied to every seat simultaneously:

>    If a seat is empty (L) and there are no occupied seats adjacent to it, the seat becomes occupied.
>    If a seat is occupied (#) and four or more seats adjacent to it are also occupied, the seat becomes empty.
>    Otherwise, the seat's state does not change.

--- Data ---
```{r eval = FALSE}
input_11 <- read_lines(here("2020","day-11.txt"))
```

--- Cleaning ---

```{r eval = FALSE}
seat_map <- tibble(col = input_11) %>% 
  mutate(col = str_split(col,"")) %>% 
  unnest_wider(col,names_sep = "_") %>% 
  mutate_all(~replace(.x,.x == ".",NA)) %>% 
  mutate_all(~replace(.x,.x == "L",0)) %>% 
  mutate_all(as.numeric)

seat_matrix <- as.matrix(seat_map)

seat_index <- seat_map %>% 
  rownames_to_column(var = "row_index") %>% 
  mutate(row_index = as.numeric(row_index)) %>% 
  pivot_longer(cols = -row_index,names_to = "column_index") %>% 
  mutate(column_index = parse_number(column_index))
```

Functions

```{r eval = FALSE}
generate_adjacent <- function(row_index,column_index,seat_matrix){
  
  check <- seat_matrix[[row_index,column_index]]
  
  if(is.na(check)) return(NA)
  
  r <- c(row_index - 1, row_index, row_index + 1)
  c <- c(column_index -1, column_index, column_index + 1)
  
  x <- expand_grid(row = r,col = c) %>% 
    filter(!(row==row_index & col==column_index)) %>% 
    filter(between(row,1,nrow(seat_matrix)),between(col,1,ncol(seat_matrix))) %>% 
    mutate(value = map2_dbl(row,col,~seat_matrix[[.x,.y]])) %>% 
    filter(!is.na(value))
    
  return(x)
}

get_seat <- function(row,col,p1){
  p1[[row, col]]
}

count_occupied <- function(adjacent_seats,p1){

  if(is.logical(adjacent_seats)) return(NA)
  
  x <- adjacent_seats %>% 
    mutate(value = map2_dbl(.data$row,.data$col,get_seat,p1))
  
  sum(x$value,na.rm = TRUE)
}

```

--- Problem 1 ---

> Simulate your seating area by applying the seating rules repeatedly until no seats change state. How many seats end up occupied?

```{r eval = FALSE}
p1_index <- seat_index %>% 
  mutate(adjacent_seats = future_map2(row_index,column_index,generate_adjacent,seat_matrix))

p1 <- seat_matrix
total_value <- sum(p1,na.rm = TRUE)
total_change <- 1
iteration <- 1
list_changes <- c()

while(total_change!=0){
  
  p1_index <- p1_index %>% 
    mutate(adjacent_status = future_map_dbl(adjacent_seats,count_occupied,p1),
           value = case_when(adjacent_status == 0 ~ 1,
                              adjacent_status >=4 ~ 0,
                              TRUE ~ value))
  
  p1 <- matrix(p1_index$value,nrow = nrow(seat_matrix), ncol = ncol(seat_matrix), byrow=TRUE)
  
  change <- sum(p1,na.rm = TRUE)
  total_change <- change - total_value
  total_value <- change
  
  list_changes <- c(list_changes,change)
  
  iteration <- iteration + 1
  
  print(iteration)
  
}
```
```{r include = FALSE}
change <- 2243
```
```{r}
change # last change
```

--- Problem 2 ---

> People don't just care about adjacent seats - they care about the first seat they can see in each of those eight directions!

So rewrite the adjacent seats function first, then rerun the loop and get seats etc. Trying matrices for faster speed.

Functions

```{r eval = FALSE}
gen_los <- function(row_dir,col_dir,row_index,col_index,row_max,col_max,seat_matrix){
  if(!between(row_index + row_dir,1,row_max)) return(NULL)
  if(!between(col_index + col_dir,1,col_max)) return(NULL)
  
  value <- seat_matrix[[row_index + row_dir,col_index + col_dir]]
  
  if(!is.na(value)) return(list(row = row_index + row_dir, col = col_index + col_dir))
  
  new_rowindex <- row_index + row_dir
  new_colindex <- col_index + col_dir
  
  gen_los(row_dir,col_dir,new_rowindex,new_colindex,row_max,col_max,seat_matrix)
}

generate_los <- function(row_index,column_index,seat_matrix){
  
  check <- seat_matrix[row_index,column_index]
  
  if(is.na(check)) return(NA)
  
  r <- c(-1, 0, 1)
  c <- c(-1, 0, 1)
  
  x <- expand_grid(row = r,col = c) %>% 
    filter(!(row == 0 & col == 0))
  
  y <- map2_dfr(x$row,x$col,gen_los,row_index,column_index,nrow(seat_matrix),ncol(seat_matrix),seat_matrix)
  
  return(y)
}

# get_seat2 <- memoise::memoise(get_seat2,cache = cache_filesystem(here("2020/day-11-cache")))

get_seat2 <- function(row,col,iteration){
  p2[row,col]
}

count_occupied2 <- function(adjacent_seats,iteration){
  
  if(is.logical(adjacent_seats)) return(NA)
  
  x <- map2_dbl(adjacent_seats$row,adjacent_seats$col,get_seat2,iteration) 
  
  sum(x,na.rm = TRUE)
}
```


```{r eval = FALSE}
p2_index <- seat_index %>% 
  mutate(row_index = as.numeric(row_index),
         adjacent_seats = future_map2(row_index,column_index,generate_los,seat_matrix))
```

```{r eval = FALSE}
p2 <- seat_matrix
total_value2 <- sum(p2,na.rm = TRUE)
total_change2 <- 1
iteration2 <- 1
list_changes2 <- c()

while(total_change2!=0){
  
  p2_index <- p2_index %>% 
    mutate(adjacent_status = future_map_dbl(adjacent_seats,count_occupied2,iteration2),
           value = case_when(adjacent_status == 0 ~ 1,
                             adjacent_status >=5 ~ 0,
                             TRUE ~ value))
  
  p2 <- matrix(p2_index$value,nrow = nrow(seat_matrix), ncol = ncol(seat_matrix), byrow=TRUE)
  
  change2 <- sum(p2,na.rm = TRUE)
  total_change2 <- change2 - total_value2
  total_value2 <- change2
  
  list_changes2 <- c(list_changes2,change2)
  
  iteration2 <- iteration2 + 1
  
  print(iteration2)
  
}
```
```{r include = FALSE}
change2 <- 2027
```
```{r}
change2
```
