---
title: "Adv2020 - Day Fifteen"
author: "Tan Ho"
date: "2020-12-15"
output: html_document
---

Day fifteen - Rambunctious Recitation!

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(slider)

  options(scipen  =  9999999)
  knitr::opts_chunk$set(echo = TRUE)
})
```

--- Description ---

> So, after the starting numbers, each turn results in that player speaking aloud either 0 (if the last number is new) or an age (if the last number is a repeat).

--- Data ---

```{r eval = FALSE}
input_15 <- c(15,12,0,14,3,1)
```

--- Problem 1 ---

```{r eval = FALSE}
p1 <- tibble(index = 1:6,value = input_15) %>% 
  bind_rows(list(index = 7:2020))

find_next <- function(i){
  prev <- p1$value[i - 1]
  new <- 0
  
  if(prev %in% p1$value[-(i-1)]) {
    new <- p1 %>% 
      filter(prev == value) %>% 
      slice_tail(n = 2) %>% 
      pull(index) %>% 
      diff()
    }
  
  p1$value[i] <<- new
}

walk(7:2020,find_next)
tail(p1$value,1)
```
```{r echo = FALSE}
249
```

--- Problem 2 ---

30,000,000 iterations slows this down heck of a lot! 

Listing changes I made to first solution: 

- instead of storing what each iteration said, store when each number was last said
- reconfigure to store a vector of previous value and a second vector of value before previous value
- pre-emptively build out the previous vectors to 100,000 long - if overflow, increase incrementally - this avoids copy on modify
- pass solutions from prev_1 to prev_2 by location
- don't recalculate indexes twice for each loop - pre-set it before the loop and then do it at the end of loop
- superassigning is slow, move the superassigned thing inside the function
- tibbles are slow. matrices are slow. data.table is slow in this context too. 

```{r eval = FALSE}

find_next2 <- function(p1, range) {
  
  p2 <- p1 %>% 
    group_by(value) %>% 
    slice_tail(n=2) %>%
    mutate(rank = rank(desc(index))) %>% 
    ungroup() %>% 
    pivot_wider(names_from = rank, names_prefix = "prev_",values_from = index) %>% 
    arrange(value) %>% 
    left_join(
      y = .,
      x = tibble(value = 0:100000),
      by = 'value'
    )
  
  value <- p2$value %>% as.integer()
  
  prev_1 <- p2$prev_1
  
  prev_2 <- p2$prev_2
  
  index <- which(prev_1 == max(prev_1,na.rm=TRUE))
  
  for (i in range) {
    
    prev_value <- prev_2[index]
    index <- 1
    
    if(!is.na(prev_value)) index <- i - prev_value
    
    prev_2[index] <- prev_1[index]
    prev_1[index] <- i
    
    # if(i %% 1000000 ==0) message(i)
  }
  
  # Return the last value
  return(index - 1)
}
tictoc::tic()
find_next2(p1,2021:30000000)
tictoc::toc()
```
```{r echo = FALSE}
41687
message("11.85 sec elapsed")
```

