---
title: "Adv2020 - Day Seven"
author: "Tan Ho"
date: "2020-12-07"
output: html_document
---

Day seven! "Handy Haversacks"

```{r setup, message = FALSE}
library(tidyverse)
library(here)

knitr::opts_chunk$set(echo = TRUE)
```

> Due to recent aviation regulations, many rules (your puzzle input) are being enforced about bags and their contents; bags must be color-coded and must contain specific quantities of other color-coded bags. Apparently, nobody responsible for these regulations considered how long they would take to enforce!
>
> For example, consider the following rules:
>
> - light red bags contain 1 bright white bag, 2 muted yellow bags.
> - dark orange bags contain 3 bright white bags, 4 muted yellow bags.
> - bright white bags contain 1 shiny gold bag.
> - muted yellow bags contain 2 shiny gold bags, 9 faded blue bags.
> - shiny gold bags contain 1 dark olive bag, 2 vibrant plum bags.
> - dark olive bags contain 3 faded blue bags, 4 dotted black bags.
> - vibrant plum bags contain 5 faded blue bags, 6 dotted black bags.
> - faded blue bags contain no other bags.
> - dotted black bags contain no other bags.
>
> These rules specify the required contents for 9 bag types. In this example, every faded blue bag is empty, every vibrant plum bag contains 11 bags (5 faded blue and 6 dotted black), and so on.
>
> You have a shiny gold bag. If you wanted to carry it in at least one other bag, how many different bag colors would be valid for the outermost bag? (In other words: how many colors can, eventually, contain at least one shiny gold bag?)
>
>In the above rules, the following options would be available to you:
>
>    - A bright white bag, which can hold your shiny gold bag directly.
>    - A muted yellow bag, which can hold your shiny gold bag directly, plus some other bags.
>    - A dark orange bag, which can hold bright white and muted yellow bags, either of which could then hold your shiny gold bag.
>    - A light red bag, which can hold bright white and muted yellow bags, either of which could then hold your shiny gold bag.
>
> So, in this example, the number of bag colors that can eventually contain at least one shiny gold bag is 4.
>
> How many bag colors can eventually contain at least one shiny gold bag? 


### Data
```{r}
input_07 <- read_lines(here("2020","day-07.txt"))
```

### Cleaning

```{r}

df_rules <- tibble(input = input_07) %>% 
  separate(input,into = c("name","rule"),sep = " bags contain ") %>% 
  mutate(rule = str_split(rule,"bags|bag")) %>% 
  unnest_longer(rule)  %>% 
  mutate(rule = str_remove(rule,","),
         rule = str_remove(rule,"\\."),
         rule = str_squish(rule)) %>% 
  filter(rule != "") %>% 
  mutate(rule_qty = parse_number(rule,na = "no other"),
         rule_colour = if_else(is.na(rule_qty), 
                               NA_character_,
                               str_remove(rule,as.character(rule_qty))),
         rule_colour = str_squish(rule_colour))

```

### Problem 1 - how many bag colours eventually contain one shiny gold bag?

```{r}

find_parents <- function(colour,df_rules){
  
  x <- df_rules %>% 
    filter(rule_colour == colour, rule_qty >= 1)
  
  map_dfr(x$name,find_parents,df_rules) %>% 
    bind_rows(x,.)
  
}

parents <- find_parents("shiny gold",df_rules)

unique(parents$name) %>% length()

```

### Problem 2 - how many bags are inside your shiny gold bag?

```{r}

library(furrr)

plan(multiprocess)

find_children <- function(colour,df_rules){
  
  x <- df_rules %>% 
    filter(name == colour,!is.na(rule_qty))
  
  if(nrow(x)==0) return(tibble())
  
  x <- x %>% 
    mutate(rule_colour = map2(rule_colour,rule_qty,rep_len)) %>% 
    unnest_longer(rule_colour)
  
  y <- future_map_dfr(x$rule_colour,find_children,df_rules)
  
  bind_rows(x,y)
}

children <- find_children("shiny gold",df_rules)

nrow(children)

```

