---
title: "Adv2020 - Day Three"
author: "Tan Ho"
date: "2020-12-03"
output: html_document
---

Day three!

```{r message = FALSE}
library(tidyverse)
library(here)

knitr::opts_chunk$set(echo = TRUE)
```

### Problem

> You start on the open square (.) in the top-left corner and need to reach the bottom (below the bottom-most row on your map).
> 
> The toboggan can only follow a few specific slopes (you opted for a cheaper model that prefers rational numbers); start by counting all the trees you would encounter for the slope right 3, down 1:
> 
> From your starting position at the top-left, check the position that is right 3 and down 1. Then, check the position that is right 3 and down 1 from there, and so on until you go past the bottom of the map.
>
> Starting at the top-left corner of your map and following a slope of right 3 and down 1, how many trees would you encounter?

### Input Data

Converting to a matrix of 1s and 0s, where "#" is 1 and "." is 0

```{r}
input_matrix <- read_table("day-03.txt",col_names = "x") %>% 
  mutate(x = str_split(x,"")) %>% 
  unnest_wider(x,names_sep = "_") %>% 
  mutate_all(~case_when(.x == "#" ~ 1, .x == "." ~ 0)) %>% 
  as.matrix()
```

### Problem 1

> Starting at the top-left corner of your map and following a slope of right 3 and down 1, how many trees would you encounter?

Key here is that the pattern repeats to the right until you get down to the bottom, i.e. until you access row 323 in this case.

```{r}
input_rows <- nrow(input_matrix)

input_rows

```
So 323 is the goal rows, and that means we'll need ~ `r 323 * 3` columns. We can tackle this cbind with a while loop. 

```{r}
input_columns <- input_rows * 3

wide_matrix <- input_matrix

while(ncol(wide_matrix) < input_columns) wide_matrix <- cbind(wide_matrix,input_matrix)
```

Now building an accessor tibble with row numbers and column numbers

```{r}
row_accessor <- 2:input_rows # start at 1,1 so the next row is 2

column_accessor <- 1:input_rows * 3 + 1 # start at 1,1 and move right three, so next column is 4

column_accessor <- column_accessor[-323] # need the accessors to be same length

indices <- tibble(
  row = row_accessor,
  column = column_accessor) %>%
  mutate(path = map2_dbl(row,column,~wide_matrix[.x,.y]))
  
head(indices)

sum(indices$path)

```

### Problem 2: Find this path for multiple slopes

Find the path for

- 1,1
- 3,1
- 5,1
- 7,1
- 1,2

A good time to generalise the previous steps!

```{r}

fn_slope <- function(right,down,matrix){
  
  # Calculate number of columns
  column_count <- (nrow(matrix) * right)
  
  # Create matrix
  wide_matrix <- matrix
  while(ncol(wide_matrix) < column_count) wide_matrix <- cbind(wide_matrix,matrix)
  
  # Create accessors
  row_accessor <- (1:nrow(matrix) * down) + 1
  row_accessor <- row_accessor[row_accessor <=nrow(matrix)]
  
  column_accessor <- (1:nrow(matrix) * right) + 1
  column_accessor <- column_accessor[1:length(row_accessor)]
  
  x <- tibble(row = row_accessor,
              column = column_accessor) %>% 
    mutate(path = map2_dbl(row,column,~wide_matrix[.x,.y]))
  
  sum(x$path)
}

toboggan_paths <- tibble(right = c(1,3,5,7,1),
                         down = c(1,1,1,1,2)) %>% 
  mutate(paths = map2_dbl(right,down,fn_slope,input_matrix))

toboggan_paths

reduce(toboggan_paths$paths,`*`)

```
