---
title: "Adv2020 - Day Twelve"
author: "Tan Ho"
date: "2020-12-12"
output: html_document
---

Day twelve - Rain Risk!

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  # library(slider)
  # library(furrr)
  # library(memoise)
  
  # plan(multisession)
  
  knitr::opts_chunk$set(echo = TRUE)
})
```

--- Description ---
>The navigation instructions (your puzzle input) consists of a sequence of single-character actions paired with integer input values. After staring at them for a few minutes, you work out what they probably mean:
>
>    - Action N means to move north by the given value.
>    - Action S means to move south by the given value.
>    - Action E means to move east by the given value.
>    - Action W means to move west by the given value.
>    - Action L means to turn left the given number of degrees.
>    - Action R means to turn right the given number of degrees.
>    - Action F means to move forward by the given value in the direction the ship is currently facing.

--- Data --- 
```{r eval = FALSE}
input_12 <- read_lines(here("2020","day-12.txt"))
```

--- Cleaning ---

```{r eval = FALSE}
instructions <- tibble(instruction = input_12) %>%
  extract(instruction,into = c("instruction","num"),regex = "^([A-z])([0-9]+)",convert = TRUE)

```

--- Problem 1 ---

> Figure out where the navigation instructions lead. What is the Manhattan distance between that location and the ship's starting position?

Setting out some functions.

```{r eval = FALSE}
move_nesw1 <- function(direction,value,current_position){
  
  switch(direction,
         "N" = current_position$y <- current_position$y + value,
         "E" = current_position$x <- current_position$x + value,
         "S" = current_position$y <- current_position$y - value,
         "W" = current_position$x <- current_position$x - value
         )
  return(current_position)
}

move_lr1 <- function(direction,value,current_position){
  switch(direction,
         "L" = current_position$direction <- current_position$direction + value,
         "R" = current_position$direction <- current_position$direction - value)
  
  if(current_position$direction >= 360) current_position$direction <- current_position$direction %% 360
  
  if(current_position$direction < 0) current_position$direction <- current_position$direction + 360
  
  return(current_position)
}

move_fb1 <- function(value,current_position){
  
  dir <- current_position$direction %>% as.character 
  
  switch (dir,
    '0' = current_position$x <- current_position$x + value,
    '90' = current_position$y <- current_position$y + value,
    '180' = current_position$x <- current_position$x - value,
    '270' = current_position$y <- current_position$y - value
  )
  
  return(current_position)
}

decide_move1 <- function(current_position,direction,value){
  
  switch(direction,
         "L" = ,
         "R" = move_lr1(direction,value,current_position),
         "F" = move_fb1(value,current_position),
         "N" = ,
         "E" = ,
         "S" = ,
         "W" = move_nesw1(direction,value,current_position))
}
```

Now run through the instructions:

```{r eval = FALSE}
current_position <- list(
  direction = 0, # where 0 = East
  x = 0,
  y = 0
)
positions <- current_position

for(i in seq_len(nrow(instructions))){

  new_position <- decide_move1(current_position,instructions$instruction[[i]],instructions$num[[i]])
  
  positions <- bind_rows(positions,new_position)
  
  current_position <- new_position
}
```
```{r include = FALSE}
current_position <- list(direction = 90, x = -397, y = -399)
```
```{r}
current_position

abs(current_position$x) + abs(current_position$y)

```

--- Problem 2 ---

> Figure out where the navigation instructions actually lead. What is the Manhattan distance between that location and the ship's starting position?

Rewriting previous functions:

```{r eval = FALSE}
move_waypoint_nesw <- function(direction,value,current_position){
  
  switch(direction,
         "N" = current_position$waypoint_y <- current_position$waypoint_y + value,
         "E" = current_position$waypoint_x <- current_position$waypoint_x + value,
         "S" = current_position$waypoint_y <- current_position$waypoint_y - value,
         "W" = current_position$waypoint_x <- current_position$waypoint_x - value
         )
  return(current_position)
}

move_waypoint_lr <- function(direction,value,current_position){

  new_position <- current_position
  
  if(value == 90){
    if(direction == "L"){
      new_position$waypoint_x <- current_position$waypoint_y * -1
      new_position$waypoint_y <- current_position$waypoint_x
    }
    if(direction == "R"){
      new_position$waypoint_x <- current_position$waypoint_y 
      new_position$waypoint_y <- current_position$waypoint_x * -1
    }
  }
  
  if(value == 180){
    new_position$waypoint_x <- current_position$waypoint_x * -1
    new_position$waypoint_y <- current_position$waypoint_y * -1
  }
  
  if(value == 270){
    if(direction == "R"){
      new_position$waypoint_x <- current_position$waypoint_y * -1
      new_position$waypoint_y <- current_position$waypoint_x
    }
    if(direction == "L"){
      new_position$waypoint_x <- current_position$waypoint_y 
      new_position$waypoint_y <- current_position$waypoint_x * -1
    }
  }
  
  return(new_position)
}

move_fb_waypoint <- function(value,current_position){
  
  dir <- current_position$direction %>% as.character 
  
  current_position$ship_x <- current_position$ship_x + (value * current_position$waypoint_x)
  current_position$ship_y <- current_position$ship_y + (value * current_position$waypoint_y)
  
  return(current_position)
}

decide_move2 <- function(current_position,direction,value){
  
  switch(direction,
         "L" = ,
         "R" = move_waypoint_lr(direction,value,current_position),
         "N" = ,
         "E" = ,
         "S" = ,
         "W" = move_waypoint_nesw(direction,value,current_position),
         "F" = move_fb_waypoint(value,current_position)
  )
}
```

```{r eval = FALSE}
current_position <- list(
  waypoint_x = 10,
  waypoint_y = 1,
  ship_x = 0,
  ship_y = 0
)

positions <- current_position

for(i in seq_len(nrow(instructions))){

  new_position <- decide_move2(current_position,instructions$instruction[[i]],instructions$num[[i]])
  
  positions <- bind_rows(positions,new_position)
  
  current_position <- new_position
}
```
```{r include = FALSE}
current_position <- list(waypoint_x = 26, waypoint_y = 37, ship_x = -24417, ship_y = -15029)
```
```{r}
current_position

abs(current_position$ship_x) + abs(current_position$ship_y)
```
