---
title: "Adv2020 - Day Twenty-One"
author: "Tan Ho"
date: "2020-12-21"
output: html_document
---

Day twenty-one - Allergen Assessment!

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(glue)

  knitr::opts_chunk$set(echo = TRUE)
})

  options(scipen = 9999999)
  options(dplyr.summarise.inform = FALSE)
```

--- Description ---

--- Data ---
```{r}
input_21 <- read_lines(here("2020", "day-21.txt"))

input_21_e <- read_lines(here("2020","day-21-e.txt"))

```

--- Cleaning ---

```{r}
ingredient_list <- tibble(i = input_21) %>% 
  mutate(id = row_number()) %>% 
  separate(i, into = c("ingredient","allergen"), sep = " \\(contains ") %>% 
  separate_rows(ingredient,sep = " ") %>%
  mutate(allergen = str_remove(allergen,"\\)")) %>% 
  separate_rows(allergen,sep = ", ")
```

--- Problem 1 ---

Count non-allergens. Spent some time thinking about this one, but basically I'm grouping by allergen and getting the ingredients that occur the most for each allergen. If it's a single ingredient, that's the one (and then eliminates that ingredient from being the allergen from any of the others). Loop through this to create a table of allergens. 

```{r}
top_allergens <- ingredient_list %>% 
  group_by(ingredient,allergen) %>% 
  summarise(n = n())%>% 
  group_by(allergen) %>% 
  slice_max(n) %>% 
  mutate(nn = n()) %>%  
  arrange(nn,desc(n),allergen)

allergen_options <- top_allergens

allergen_df <- tibble()

while(nrow(allergen_options)>0) {
  
  x <- allergen_options %>% 
    filter(nn == 1) %>% 
    select(ingredient, allergen)

  allergen_df <- bind_rows(allergen_df, x)
  
  allergen_options <- allergen_options %>% 
    filter(!ingredient %in% allergen_df$ingredient, !allergen %in% allergen_df$allergen) %>% 
    group_by(allergen) %>% 
    mutate(nn = n())
}

non_allergens <- ingredient_list %>% 
  filter(!ingredient %in% allergen_df$ingredient) %>% 
  distinct(ingredient,id)

nrow(non_allergens)
```

--- Problem 2 ---

...is only to arrange the allergens in order? what?

```{r}

allergen_df %>% 
  arrange(allergen) %>% 
  pull(ingredient) %>% 
  paste(collapse = ",")

```
...that was ... easy?!
