---
title: "Adv2020 - Day Seventeen"
author: "Tan Ho"
date: "2020-12-17"
output: html_document
---

Day seventeen - Conway Cubes!

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(furrr)
  plan(multisession)

  options(scipen = 9999999)
  knitr::opts_chunk$set(echo = TRUE)
})
```


--- Description ---

> The experimental energy source is based on cutting-edge technology: a set of Conway Cubes contained in a pocket dimension! When you hear it's having problems, you can't help but agree to take a look.
> 
> The pocket dimension contains an infinite 3-dimensional grid. At every integer 3-dimensional coordinate (x,y,z), there exists a single cube which is either active or inactive.

So essentially Conway's game of life again...but in three dimensions?

--- Data ---

```{r eval = FALSE}
input_17 <- read_lines(here("2020", "day-17.txt"))
input_17_e <- read_lines(here("2020", "day-17-example1.txt"))
```

--- Cleaning ---

Exploring a new part of R to me: arrays, which are vertically stacked matrices.

```{r eval = FALSE}
example_matrix <- tibble(x = input_17_e) %>%
  mutate(x = str_split(x, "")) %>%
  unnest_wider(x, names_sep = "_") %>%
  mutate_all(~ case_when(.x == "#" ~ 1, TRUE ~ 0)) %>%
  as.matrix()

example_empty <- matrix(
  data = rep(0, length(example_matrix)),
  nrow = nrow(example_matrix),
  ncol = ncol(example_matrix)
)

example_array <- c(example_empty, example_matrix, example_empty) %>%
  array(dim = c(nrow(example_matrix), ncol(example_matrix), 3))

input_matrix <- tibble(x = input_17) %>%
  mutate(x = str_split(x, "")) %>%
  unnest_wider(x, names_sep = "_") %>%
  mutate_all(~ case_when(.x == "#" ~ 1, TRUE ~ 0)) %>%
  as.matrix()

input_empty <- matrix(
  data = rep(0, length(input_matrix)), byrow = TRUE,
  nrow = nrow(input_matrix),
  ncol = ncol(input_matrix)
)

input_array <- c(input_empty, input_matrix, input_empty) %>%
  array(dim = c(nrow(input_matrix), ncol(input_matrix), 3))
```
```{r echo = FALSE}
example_array <- structure(c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 1, 0, 1, 
1, 0, 0, 0, 0, 0, 0, 0, 0, 0), .Dim = c(3L, 3L, 3L))
```

Okay! So now we have the starting states of the example as `example_array` and the actual problem as `input_array`. We can access the dimensions of the arrays via 

```{r}
dim(example_array)
```

and access a specific layer with array[row,column,layer].

```{r}
example_array[1, 2, 2]
```

--- Problem 1 ---

We can build out a list of indices with the dimensions, I think!

```{r eval = FALSE}
# array <- example_array

get_indices <- function(array) {
  dimensions <- dim(array)

  actual_indices <- crossing(
    x = seq_len(dimensions[[1]]),
    y = seq_len(dimensions[[2]]),
    z = seq_len(dimensions[[3]])
  ) %>%
    arrange(z, x, y) %>%
    mutate(
      value = pmap_dbl(list(x, y, z), ~ array[...]),
      x = x + 1,
      y = y + 1
    )

  indices <- crossing(
    x = seq_len(dimensions[[1]] + 2),
    y = seq_len(dimensions[[2]] + 2),
    z = seq_len(dimensions[[3]])
  ) %>%
    arrange(z, x, y) %>%
    left_join(actual_indices, by = c("x", "y", "z")) %>%
    mutate_all(replace_na, 0)

  return(indices)
}

example_indices <- get_indices(example_array)
```

Now to create a function to check the neighbouring indices and count nearby occupied cells.

```{r eval = FALSE}

# indices <- example_indices
# x <- 1
# y <- 2
# z <- 2

count_neighbours <- function(x, y, z, indices) {
  x_max <- max(indices$x)
  y_max <- max(indices$y)
  z_max <- max(indices$z)

  increments <- c(-1, 0, 1)

  neighbours <- crossing(
    x = increments + x,
    y = increments + y,
    z = increments + z
  ) %>%
    filter(
      !(.data$x == .env$x & .data$y == .env$y & .data$z == .env$z),
      between(.data$x, 1, x_max),
      between(.data$y, 1, y_max),
      between(.data$z, 1, z_max)
    ) %>%
    mutate(value = pmap_dbl(list(x, y, z), ~ indices$value[indices$x == ..1 & indices$y == ..2 & indices$z == ..3])) %>%
    pull(value) %>%
    sum()

  return(neighbours)
}

example_neighbours <- example_indices %>%
  mutate(neighbours = pmap_dbl(list(x, y, z), count_neighbours, .))
```

Now that we have the neighbours, we need to increment their values based on the rules. 

```{r eval = FALSE}
# indices <- example_neighbours

apply_rules <- function(indices) {
  x <- indices %>%
    mutate(new_value = case_when(
      value == 1 & neighbours %in% c(2, 3) ~ 1,
      value == 1 ~ 0,
      value == 0 & neighbours == 3 ~ 1,
      TRUE ~ 0
    ))
  return(x)
}

example_count <- example_neighbours %>%
  apply_rules()
```

(I was stumped for a while here as to why my puzzle input did not match up to the example, but consulting no-spoilers-reddit seems that they're dropping empty x/y dimensions which is quite annoying!)

Now to increment the array in every direction (but sticking with a list of indices for now, I find that easier to work with mentally)

```{r eval = FALSE}
# indices <- example_count
grow_indices <- function(indices) {
  new_x <- max(indices$x + 2) %>% seq_len()
  new_y <- max(indices$y + 2) %>% seq_len()
  new_z <- max(indices$z + 2) %>% seq_len()

  new_indices <- indices %>%
    transmute(
      x = x + 1,
      y = y + 1,
      z = z + 1,
      value = new_value
    )

  indices <- crossing(
    x = new_x,
    y = new_y,
    z = new_z
  ) %>%
    left_join(new_indices, by = c("x", "y", "z")) %>%
    mutate(value = replace_na(value, 0))

  return(indices)
}
# grow_indices(example_count)
```
Wrap it all into a caller function and for-loop.

```{r eval = FALSE}

# array <- example_array
# times <- 6

run_conwaycube <- function(array, times) {
  indices <- get_indices(array)

  for (i in seq_len(times)) {
    indices <- indices %>%
      mutate(neighbours = pmap_dbl(list(x, y, z), count_neighbours, .)) %>%
      apply_rules() %>%
      grow_indices()

    message(paste(i, Sys.time()))
  }

  return(sum(indices$value))
}

run_conwaycube(input_array,6)
```
```
1 2020-12-17 10:26:13
2 2020-12-17 10:26:16
3 2020-12-17 10:26:22
4 2020-12-17 10:26:33
5 2020-12-17 10:26:52
6 2020-12-17 10:27:23

280
```

--- Problem 2 ---

FOUR DIMENSIONS?! (brain explodes)

Actually, maybe I can just adjust my functions for a fourth dimension `w`

```{r eval = FALSE}

get_indices_4d <- function(array) {
  
  dimensions <- dim(array)
  
  actual_indices <- crossing(
    x = seq_len(dimensions[[1]]),
    y = seq_len(dimensions[[2]]),
    z = seq_len(dimensions[[3]])
  ) %>%
    arrange(z, x, y) %>%
    mutate(
      value = pmap_dbl(list(x, y, z), ~ array[...]),
      x = x + 1,
      y = y + 1,
      w = 2
    )
  
  indices <- crossing(
    x = seq_len(dimensions[[1]] + 2),
    y = seq_len(dimensions[[2]] + 2),
    z = seq_len(dimensions[[3]]),
    w = seq_len(dimensions[[3]])
  ) %>%
    arrange(z, x, y, w) %>%
    left_join(actual_indices, by = c("x", "y", "z", "w")) %>%
    mutate_all(replace_na, 0)
  
  return(indices)
}

count_neighbours_4d <- function(x, y, z, w, indices) {
  x_max <- max(indices$x)
  y_max <- max(indices$y)
  z_max <- max(indices$z)
  w_max <- max(indices$w)
  
  increments <- c(-1, 0, 1)
  
  neighbours <- crossing(
    x = increments + x,
    y = increments + y,
    z = increments + z,
    w = increments + w
  ) %>%
    filter(
      !(.data$x == .env$x & 
          .data$y == .env$y & 
          .data$z == .env$z & 
          .data$w == .env$w),
      between(.data$x, 1, x_max),
      between(.data$y, 1, y_max),
      between(.data$z, 1, z_max),
      between(.data$w, 1, w_max)
    ) %>%
    mutate(
      value = pmap_dbl(
        list(x, y, z, w), 
        ~ indices$value[indices$x == ..1 & 
                          indices$y == ..2 & 
                          indices$z == ..3 & 
                          indices$w == ..4])) %>%
    pull(value) %>%
    sum()
  
  return(neighbours)
}

# apply rules doesn't change since rules are the same

grow_indices_4d <- function(indices) {
  new_x <- max(indices$x + 2) %>% seq_len()
  new_y <- max(indices$y + 2) %>% seq_len()
  new_z <- max(indices$z + 2) %>% seq_len()
  new_w <- max(indices$w + 2) %>% seq_len()
  
  new_indices <- indices %>%
    transmute(
      x = x + 1,
      y = y + 1,
      z = z + 1,
      w = w + 1,
      value = new_value
    )
  
  indices <- crossing(
    x = new_x,
    y = new_y,
    z = new_z,
    w = new_w
  ) %>%
    left_join(new_indices, by = c("x", "y", "z", "w")) %>%
    mutate(value = replace_na(value, 0))
  
  return(indices)
}

run_conwaycube_4d <- function(array, times) {
  
  indices <- get_indices_4d(array)
  
  message(paste(Sys.time(), nrow(indices)))
  
  for (i in seq_len(times)) {
    indices <- indices %>%
      mutate(neighbours = pmap_dbl(list(x, y, z, w), count_neighbours_4d, .)) %>%
      apply_rules() %>%
      grow_indices_4d()
    message(paste(i, Sys.time(), nrow(indices)))
  }
  return(sum(indices$value))
}

run_conwaycube_4d(input_array, 6)
```
```
2020-12-17 02:01:36 900
1 2020-12-17 02:01:41 3600
2 2020-12-17 02:02:12 9604
3 2020-12-17 02:04:46 20736
4 2020-12-17 02:15:10 39204
5 2020-12-17 02:50:04 67600
6 2020-12-17 04:35:08 108900

[1] 1696
```

Admittedly, I plugged this in and went to sleep - but it was the correct answer!

--- Iterating on Problem 2 for speed ---

Problem 2 takes quite some time to run. Trying a few approaches to trimming it down:

Firstly, as Liam Y suggested in the R4DS Slack - arrange the neighbour function so that it's only checking neighbours that have positive integer values:

```{r eval = FALSE}
count_neighbours_4d <- function(indices){

  filtered_indices <- indices %>%
    filter(value == 1) # This should reduce the amount of iteration needed!

  indices %>%
    mutate(neighbours = pmap_dbl(list(x,y,z,w),check_each_neighbour,filtered_indices))
}

check_each_neighbour <- function(x, y, z, w, indices){

  increments <- c(-1, 0, 1)

  n <- crossing(
    x = increments + x,
    y = increments + y,
    z = increments + z,
    w = increments + w
  ) %>%
    filter(
      !(.data$x == .env$x & 
          .data$y == .env$y & 
          .data$z == .env$z & 
          .data$w == .env$w)
    ) %>%
    mutate(
      value = pmap(list(.data$x,.data$y,.data$z,.data$w),
                   ~(indices$value[indices$x == ..1 & indices$y == ..2 & indices$z == ..3 & indices$w == ..4]))) %>%
    pull(value) %>%
    unlist() %>%
    sum(na.rm = TRUE)

  return(n)
}

run_conwaycube_4d <- function(array, times) {
  indices <- get_indices_4d(array)

  message(paste(Sys.time(), nrow(indices)))

  for (i in seq_len(times)) {
    indices <- indices %>%
      count_neighbours_4d() %>% 
      apply_rules() %>%
      grow_indices_4d()

    message(paste(i, Sys.time(), nrow(indices)))
  }

  return(sum(indices$value))
}

run_conwaycube_4d(input_array, 6)
```

```
2020-12-17 16:08:14 900
1 2020-12-17 16:08:18 3600
2 2020-12-17 16:08:38 9604
3 2020-12-17 16:09:29 20736
4 2020-12-17 16:11:39 39204
5 2020-12-17 16:15:26 67600
6 2020-12-17 16:25:28 108900

[1] 1696
```

Just doing that much shrunk the run-time from 2 hours 35 minutes to ~17 minutes! I can do better though, I think - in Day 11's final approach, I arranged things into a matrix and approached it from there - I think I can definitely do the same here! (and also, add parallel processing because duh)

```{r eval = FALSE}
count_neighbours_4d <- function(indices){

  v <- indices %>% 
    arrange(w,z,y,x) %>% 
    pull(value)
  
  a <- array(v,dim = c(max(indices$x),
                       max(indices$y),
                       max(indices$z),
                       max(indices$w)))

  i <- indices %>%
    mutate(neighbours = future_pmap_dbl(.l = list(.data$x,.data$y,.data$z,.data$w),.f = check_each_neighbour,a = .env$a))

  return(i)
}

check_each_neighbour <- function(x, y, z, w, a){

  increments <- c(-1, 0, 1)

  # browser()
  
  n <- crossing(
    x = increments + x,
    y = increments + y,
    z = increments + z,
    w = increments + w
  ) %>%
    filter(
      !(.data$x == .env$x & 
          .data$y == .env$y & 
          .data$z == .env$z & 
          .data$w == .env$w),
      between(.data$x,1,dim(a)[[1]]),
      between(.data$y,1,dim(a)[[2]]),
      between(.data$z,1,dim(a)[[3]]),
      between(.data$w,1,dim(a)[[4]]),
    ) %>%
    mutate(
      value = pmap_dbl(
        list(.data$x,.data$y,.data$z,.data$w),
        ~a[..1,..2,..3,..4])) %>%
    pull(value) %>%
    unlist() %>%
    sum(na.rm = TRUE)

  return(n)
}

run_conwaycube_4d <- function(array, times) {
  indices <- get_indices_4d(array)

  message(paste(Sys.time(), nrow(indices)))

  for (i in seq_len(times)) {
    indices <- indices %>%
      count_neighbours_4d() %>% 
      apply_rules() %>%
      grow_indices_4d()

    message(paste(i, Sys.time(), nrow(indices)))
  }

  return(sum(indices$value))
}

run_conwaycube_4d(input_array, 6)

```
```
2020-12-17 18:39:43 900
1 2020-12-17 18:39:45 3600
2 2020-12-17 18:39:48 9604
3 2020-12-17 18:39:56 20736
4 2020-12-17 18:40:13 39204
5 2020-12-17 18:40:44 67600
6 2020-12-17 18:41:39 108900
[1] 1696
```

Shaving it down to two minutes runtime is extremely satisfying!
