---
title: 'Advent Of Code: 2021-04'
author: "Tan Ho"
date: "2021-12-04"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(glue)
  
  knitr::opts_chunk$set(echo = TRUE)
})

options(scipen = 9999999)
options(dplyr.summarise.inform = FALSE)
```

--- Data ---

```{r eval = FALSE}
# tanho63/aoc.elf
aoc.elf::aoc_get(4)
```

```{r}
input <- readLines(here::here("2021/day-04-input.txt"))
```

I find it easier to parse stuff when it's in tibbles, this  happens to be the fastest way for me to lump all the boards together in my mind. Also creating a score matrix of the same length.

```{r}
bingo_calls <- input[[1]] |> strsplit(",") |> unlist() |> as.numeric()

bingo_boards <- tibble(x = input[-1]) |> 
  mutate(board_id = cumsum(x == "")) |> 
  filter(x != "") |> 
  group_by(board_id) |> 
  summarise(boards = paste(x, collapse = " ") |> str_squish()) |> 
  mutate(boards = strsplit(boards, "\\s+") |> 
           map(~as.numeric(.x) |> matrix(ncol = 5, byrow = TRUE)),
         selected = list(matrix(rep(0,25),ncol = 5))
  )

```

--- Part 1 ---

Create a function to check a bingo board for success, and a different function to call the bingo number. Then a nested loop to call all of the boards for each bingo number, breaking if any are successful

```{r}
# id_matrix <- bingo_boards$boards[[1]]
# score_matrix <- matrix(rep(0,25),ncol = 5)
# bingo_call <- 97

check_board <- function(m){
  rows <- rowSums(m,na.rm = TRUE)
  cols <- colSums(m, na.rm = TRUE)
  diagon <- sum(m[1,1],m[2,2],m[3,3],m[4,4],m[5,5])
  alley <- sum(m[5,1],m[4,2],m[3,3],m[2,4],m[1,5])
  
  if(any(rows == 5, cols == 5, diagon == 5, alley == 5)) return(TRUE)
  
  return(FALSE)
}

call_number <- function(score_matrix, bingo_call, id_matrix){
  
  l <- which(id_matrix %in% bingo_call)
  score_matrix[l] <- 1
  return(score_matrix)
}

boards <- bingo_boards$boards
scores <- bingo_boards$selected

for (n in bingo_calls){
  for (b in bingo_boards$board_id){
    scores[[b]] <- call_number(scores[[b]], bingo_call = n, id_matrix = boards[[b]])
    if(check_board(scores[[b]])) break
  }
  if(check_board(scores[[b]])) break
}

```

```{r}
sum((scores[[b]] == 0) * (boards[[b]])) * n
```

--- Part 2 ---

Same as before but storing which bingo call resulted in a success state. 

```{r}
boards2 <- bingo_boards$boards
scores2 <- bingo_boards$selected

board_status <- rep(0,100)
counter <- 1
for (n in bingo_calls){
  for (b in bingo_boards$board_id){
    if(check_board(scores2[[b]])) next
    scores2[[b]] <- call_number(scores2[[b]], bingo_call = n, id_matrix = boards2[[b]])
    if(check_board(scores2[[b]])) board_status[b] <- counter
  }
  if(all(sapply(scores2, check_board))) break
  counter <- counter + 1
}

last <- which.max(board_status)

sum((scores2[[last]] == 0) * (boards2[[last]])) * n
```

