---
title: 'Advent Of Code: 2021-03'
author: "Tan Ho"
date: "2021-12-03"
output: html_document
---

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(glue)
  
  knitr::opts_chunk$set(echo = TRUE)
})

options(scipen = 9999999)
options(dplyr.summarise.inform = FALSE)
```

--- Data ---

```{r eval = FALSE}
aoc.elf::aoc_get(3)
```

```{r}
input <- readLines(here::here("2021/day-03-input.txt"))
```

Cleaning the data to a tibble with columns for each digit, because I find that to be the most useful/workable input format

```{r}
i <- tibble(x = input) |> 
  mutate(y = strsplit(x,""),
         x = NULL) %>%
  {suppressMessages(unnest_wider(.,y))}

i
```

--- Part 1 ---

Taking the tidyverse way out here by summarising and counting...which may bite me...

```{r}
p1 <- i |> 
  mutate(x = row_number()) |> 
  pivot_longer(-x,names_to = "y") |> 
  mutate(y = str_remove(y,"...") |> as.numeric()) |> 
  group_by(y) |> 
  count(value)

head(p1)

gamma <- p1 |> 
  slice_max(n) |> 
  pull(value) |> 
  paste(collapse = "") |> 
  strtoi(base = 2L)

epsilon <- p1 |> 
  slice_min(n) |> 
  pull(value) |> 
  paste(collapse = "") |> 
  strtoi(base = 2L)

gamma * epsilon
```

--- Part 2 ---

Tidyverse bit me a bit here, going back and adding parsing functions and reducing them over each column of the matrix.

```{r}
p2 <- as.matrix(i) |> unname()

parser <- function(m, pos, most = TRUE){
  if(is.null(dim(m))) return(m)
  
  keep <- .mostleast(m[,pos], most)
  return(m[m[,pos] == keep,])
}

.mostleast <- function(v, most = TRUE){
  one <- sum(v == "1")
  zero <- sum(v == '0')
  
  if(most) return(ifelse(one >= zero, "1", "0"))
  return(ifelse(one < zero, "1", "0"))
}

oxygen <- reduce(seq_len(ncol(p2)), parser, .init = p2, most = TRUE) |> 
  paste(collapse = "") |> 
  strtoi(base = 2L)

co2 <- reduce(seq_len(ncol(p2)), parser, .init = p2, most = FALSE) |> 
  paste(collapse = "") |> 
  strtoi(base = 2L)

oxygen * co2
```

--- Part 1 redux ---

with the function built for part 2, revisiting part 1 for extra-ish credit:

```{r}

gamma2 <- map(seq_len(ncol(p2)), ~.mostleast(p2[,.x], most = TRUE)) |> 
  paste(collapse = "") |> 
  strtoi(base = 2L)
epsilon2 <- map(seq_len(ncol(p2)),~.mostleast(p2[,.x], most = FALSE)) |> 
  paste(collapse = "") |> 
  strtoi(base = 2L)

gamma2 * epsilon2
```

I can largely see why I think of the tidy/pivot way first 😂
