---
title: 'Advent Of Code: 2021-17'
author: Tan Ho
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(glue)
  
  knitr::opts_chunk$set(echo = TRUE)
})

options(scipen = 9999999)
options(dplyr.summarise.inform = FALSE)
```

--- Data ---

```{r eval = FALSE}
# tanho63/aoc.elf
aoc.elf::aoc_get(day = 17, year = 2021)
```

```{r}
input <- tibble(x = readLines(here::here("2021/day-17-input.txt"))) |> 
  extract(
    x, 
    into = c("xmin","xmax","ymin","ymax"), 
    regex = "x=([0-9,\\-]+)..([0-9,\\-]+), y=([0-9,\\-]+)..([0-9,\\-]+)",
    convert = TRUE
  )

```

--- Part 1 ---

~~solve for some kind of system of equations?~~
~~solve for inequalities?~~
~~optim?~~
guess and check lul

Highest point of Y will involve X being the first point at which X reaches zero velocity within the range.

```{r}
launch <- function(xv,yv){
  x <- 0
  y <- 0
  success <- FALSE
  max_y <- 0
  
  while(x < 227 & y > -134){
    x <- x + xv
    y <- y + yv
    xv <- (sign(xv) * -1) + xv
    yv <- yv - 1
    
    if(y > max_y) max_y <- y
    
    if(x >= 175 && x <= 227 && y >= -134 && y <= -79){
      success <- TRUE
      break
    } 
  }
  return(list(success = success, y = max_y))
}

x <- 0

repeat {
  r <- sum(seq_len(x))
  if(r >= 175 && r <= 227) break
  x <- x + 1
}

y <- 1000 # arbitrary number that I know will fail

repeat {
  y <- y-1
  res <- launch(x,y)
  if(res$success) break
  if(y < 0) break
}

res$y
```

--- Part 2 ---

Clean up launch function to return only the success, hardcode the input bounds lol

```{r}
launch <- function(xv,yv){
  x <- 0
  y <- 0
  
  success <- FALSE
  
  while(x < 227 & y > -134){
    
    x <- x + xv
    y <- y + yv
    xv <- (sign(xv) * -1) + xv
    yv <- yv - 1
    
    if(
      x >= 175 &&
      x <= 227 &&
      y >= -134 &&
      y <= -79
    ){
      success <- TRUE
      break
    } 
  }
  
  return(success)
}
```

I know: 

- x min velocity is 19 from prev step
- x max velocity is 227 (xmax)
- y min velocity is -134 (ymin) 
- y max velocity is 133 from prev step

Make a search grid and launch em all!

```{r}
g <- crossing(x = 19:227, y = -134:133)
map2_lgl(g$x,g$y,launch) |> sum()
```


