---
title: 'Advent Of Code: 2021-16'
author: Tan Ho
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
suppressPackageStartupMessages({
  library(R.utils)
  library(tidyverse)
  library(here)
  library(glue)
  
  knitr::opts_chunk$set(echo = TRUE)
})

options(scipen = 9999999)
options(dplyr.summarise.inform = FALSE)
```

--- Data ---

```{r eval = FALSE}
# tanho63/aoc.elf
aoc.elf::aoc_get(day = 16, year = 2021)
```

```{r}
input <- readLines(here::here("2021/day-16-input.txt"))
```

--- Part 1 ---

Helper functions for converting

```{r}
.hexbin <- tibble::tribble(
  ~hex, ~bin,
  "0", "0000",
   "1",  "0001",
   "2",  "0010",
   "3",  "0011",
   "4",  "0100",
   "5",  "0101",
   "6",  "0110",
   "7",  "0111",
   "8",  "1000",
   "9",  "1001",
   "A",  "1010",
   "B",  "1011",
   "C",  "1100",
   "D",  "1101",
   "E",  "1110",
   "F",  "1111"
  ) |> 
  deframe()
hex2bin <- function(hex){
  s <- strsplit(hex,"") |> unlist()
  
  paste(.hexbin[s],collapse = "")
}

bin2int <- function(bin){
  # strtoi(bin,base = 2) # no good, overflows after 2^31
  x <- as.numeric(strsplit(bin, "")[[1]])
  sum(x * 2^rev((seq_along(x)-1)))
}

```

Identify packets. 

Function to parse:

- first three characters of the binary string are the version number
- next three characters of binary identify type ID which controls how the rest of the string is parsed

if:

- type id = 4 => literal binary value
- type id != 4 => operator, also passes in the subpackets. 

functions 
- read packet version
- read packet type
- do operation for packet type

```{r}
read_packet_version <- function(packet){
  substr(packet,1,3) |> bin2int()
}

read_packet_type <- function(packet){
  substr(packet,1,3) |> bin2int()
}

read_packet_length_type <- function(packet){
  substr(packet,1,1) |> bin2int()
}

read_subpacket_bits <- function(packet){
  substr(packet,1,15) |> bin2int()
}

read_subpacket_length <- function(packet){
  substr(packet,1,11) |> bin2int()
}

read_literal_packet <- function(packet){
  
  x <- tibble(x = strsplit(packet,"") |> unlist()) |> 
    mutate(
      n = row_number() %/% 5.00001 + 1
    ) |> 
    group_by(n) |> 
    mutate(
      h = first(x) |> as.numeric(),
      x = ifelse(row_number() == 1,"",x)
    ) |> 
    group_by(n,h) |> 
    summarise(x = paste0(x, collapse = "")) |> 
    ungroup() |> 
    mutate(
      cumsum_h = cumsum(h==0),
      keep = lag(cumsum_h,default = 0) == 0
    ) |> 
    filter(keep)
    
  unprocessed <- str_sub(packet,max(x$n)*5+1)
  
  if(str_length(unprocessed) < 11) unprocessed <- ""
  
  literal_value <- x |> 
    pull(x) |> 
    paste(collapse = "") |> 
    bin2int()
  
  if(is.na(literal_value)) browser()
  
  return(list(value = literal_value, unprocessed = unprocessed))
}
```

```{r}
read_packet <- function(packet){
  original_packet <- packet
  packet_version <- read_packet_version(packet)
  packet <- str_sub(packet,4)
  packet_type <- read_packet_type(packet)
  packet <- str_sub(packet,4)
  
  ## LITERAL PACKETS (Type 4)
  if(packet_type == 4){
    # return(packet_version)
    literal <- read_literal_packet(packet)
    # For Part One
    return(
      list(version = packet_version,
           type = packet_type,
           value = literal$value,
           unprocessed = literal$unprocessed)
    )
  }
  
  packet_versions <- c(packet_version)
  packet_types <- c(packet_type)
  packet_values <- c()
  # SUBPACKETS 
  ## SUBPACKETS WITH A LENGTH CONSTRAINT
  length_type <- read_packet_length_type(packet)
  packet <- str_sub(packet,2)
  if(length_type == 0) {
    subpacket_bits <- read_subpacket_bits(packet)
    packet <- str_sub(packet,16)
    subpacket <- str_sub(packet,1,subpacket_bits)
    packet <- str_sub(packet,subpacket_bits+1)
    while(str_length(subpacket) > 10){
      returned_subpacket <- read_packet(subpacket)
      packet_versions <- c(packet_versions, returned_subpacket$version)
      packet_types <- c(packet_types, returned_subpacket$type)
      packet_values <- c(packet_values, returned_subpacket$value)
      subpacket <- returned_subpacket$unprocessed
    }
  }
  
  ## SUBPACKETS WITH A COUNT CONSTRAINT
  if(length_type == 1){
    subpacket_length <- read_subpacket_length(packet)
    packet <- str_sub(packet,12)
    for(i in seq_len(subpacket_length)){
      returned_subpacket <- read_packet(packet)
      packet_versions <- c(packet_versions, returned_subpacket$version)
      packet_types <- c(packet_types, returned_subpacket$type)
      packet_values <- c(packet_values,returned_subpacket$value)
      packet <- returned_subpacket$unprocessed
    }
  }
  
  return(list(version = packet_versions,
              original = original_packet,
              type = packet_types,
              value = packet_values,
              unprocessed = packet))
}

decode_packet <- function(hex){
  hex2bin(hex) |> read_packet()
}

```

```{r}
example <- "8A004A801A8002F478"
packet <- hex2bin(example)
packet |> read_packet()
```

Tests:
```{r}
.e1 <- decode_packet("D2FE28")
testthat::expect_equal(.e1$version |> sum(), 6)
.e2 <- decode_packet("8A004A801A8002F478")
testthat::expect_equal(.e2$version |> sum(), 16)
.e3 <- decode_packet("38006F45291200")
testthat::expect_equal(.e3$version |> sum(),9)
.e4 <- decode_packet("620080001611562C8802118E34")
testthat::expect_equal(.e4$version |> sum(),12)
.e5 <- decode_packet("C0015000016115A2E0802F182340")
testthat::expect_equal(.e5$version |> sum(),23)
.e6 <- decode_packet("A0016C880162017C3686B18A3D4780")
testthat::expect_equal(.e6$version |> sum(),31)

```

```{r}
p1 <- decode_packet(input)
sum(p1$version)
```

--- Part 2 ---

```{r}
read_packet <- function(packet){
  original_packet <- packet
  packet_version <- read_packet_version(packet)
  packet <- str_sub(packet,4)
  packet_type <- read_packet_type(packet)
  packet <- str_sub(packet,4)
  
  ## LITERAL PACKETS (Type 4)
  if(packet_type == 4){
    literal <- read_literal_packet(packet)
    return(
      list(version = packet_version,
           type = packet_type,
           value = literal$value,
           unprocessed = literal$unprocessed)
    )
  }
  subpackets <- list()
  # SUBPACKETS 
  ## SUBPACKETS WITH A LENGTH CONSTRAINT
  length_type <- read_packet_length_type(packet)
  packet <- str_sub(packet,2)
  if(length_type == 0) {
    subpacket_bits <- read_subpacket_bits(packet)
    packet <- str_sub(packet,16)
    subpacket <- str_sub(packet,1,subpacket_bits)
    packet <- str_sub(packet,subpacket_bits+1)
    while(str_length(subpacket) > 10){
      returned_subpacket <- read_packet(subpacket)
      subpackets <- c(subpackets, list(returned_subpacket))
      subpacket <- returned_subpacket$unprocessed
    }
  }
  
  ## SUBPACKETS WITH A COUNT CONSTRAINT
  if(length_type == 1){
    subpacket_length <- read_subpacket_length(packet)
    packet <- str_sub(packet,12)
    for(i in seq_len(subpacket_length)){
      returned_subpacket <- read_packet(packet)
      subpackets <- c(subpackets, list(returned_subpacket))
      packet <- returned_subpacket$unprocessed
    }
  }
  
  return(list(
    version = packet_version, 
    type = packet_type,
    packets = subpackets,
    unprocessed = packet))
}
```

```{r}
evaluate_packet <- function(packet){
  packet_type <- packet |> pluck("type")
  if(packet_type == 4) return(packet$value)
  
  subpacket_values <- c()
  
  for(i in seq_along(packet$packets)){
    subpacket_values <- c(subpacket_values, evaluate_packet(packet$packets[[i]]))
  }
  
  packet_value <- switch(
    as.character(packet_type),
    "0" = sum(subpacket_values, na.rm = TRUE),
    "1" = prod(subpacket_values, na.rm = TRUE),
    "2" = min(subpacket_values, na.rm = TRUE),
    "3" = max(subpacket_values, na.rm = TRUE),
    "5" = as.numeric(subpacket_values[1] > subpacket_values[2]),
    "6" = as.numeric(subpacket_values[1] < subpacket_values[2]),
    "7" = as.numeric(subpacket_values[1] == subpacket_values[2])
  )
  
  return(packet_value)
}
decode_packet <- function(hex){
  hex2bin(hex) |> read_packet()
}
```

```{r}
decode_packet(input) |> evaluate_packet()
```

