---
title: 'Advent Of Code: 2021-12'
author: Tan Ho
date: "`r Sys.Date()`"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(glue)
  
  knitr::opts_chunk$set(echo = TRUE)
})

options(scipen = 9999999)
options(dplyr.summarise.inform = FALSE)
```

--- Data ---

```{r eval = FALSE}
# tanho63/aoc.elf
aoc.elf::aoc_get(day = 12, year = 2021)
```

```{r}
input <- tibble(x = readLines(here::here("2021/day-12-input.txt"))) |> 
  separate(x,into = c("left","right"), sep = "\\-")

example <- c(
  "start-A",
  "start-b",
  "A-c",
  "A-b",
  "b-d",
  "A-end",
  "b-end") %>%
  tibble(x = .) |> 
  separate(x,into = c("left","right"), sep = "\\-")

example_two <- c(
  "dc-end",
"HN-start",
"start-kj",
"dc-start",
"dc-HN",
"LN-dc",
"HN-end",
"kj-sa",
"kj-HN",
"kj-dc"
)%>%
  tibble(x = .) |> 
  separate(x,into = c("left","right"), sep = "\\-")

example_three <- c(
  "fs-end",
"he-DX",
"fs-he",
"start-DX",
"pj-DX",
"end-zg",
"zg-sl",
"zg-pj",
"pj-he",
"RW-he",
"fs-DX",
"pj-RW",
"zg-RW",
"start-pj",
"he-WI",
"zg-he",
"pj-fs",
"start-RW"
)%>%
  tibble(x = .) |> 
  separate(x,into = c("left","right"), sep = "\\-")

```

--- Part 1 ---

Count all possible paths. Constraints: cannot visit small cave more than once. Must start at start and end and end. 

This should be a function that takes in a node. 
Finds the child nodes
For each child node, check to see if the child nodes are valid against:
 - have we been here before? 
 - Are we allowed to go back?
 - Is this the end?

For the valid child nodes, call the function again (recursively)

and return path on exit.


```{r eval = FALSE}

connections <- input

paths <- connections |> 
  bind_rows(connections |> 
              select(left = right, right = left))

n <- "start"
p <- ""

valid_child <- function(n,p){
  
  # if haven't been there, is ok
  if(!n %in% p) return(TRUE) 
  # if is uppercase, ok to go back
  if(toupper(n) == n) return(TRUE)
  # if not uppercase and have been there, bad
  return(FALSE)
}

pathfinder <- function(n, current_path){
  current_path <- c(current_path,n)
  
  if(n == "end") return(tibble(path = list(current_path)))
  
  child_nodes <- paths$right[paths$left == n]
  
  valid_child_nodes <- child_nodes[map_lgl(child_nodes,
                                           valid_child, 
                                           current_path)]
  
  if(length(valid_child_nodes) == 0) return(NULL)
  
  # all_paths <- map(valid_child_nodes, pathfinder, current_path)
  
  all_paths <- tibble(path = list())
  
  for(cn in valid_child_nodes) {
    all_paths <- bind_rows(all_paths,pathfinder(cn,current_path))
  }
  
  return(all_paths)
}

p1 <- pathfinder(n = "start", current_path = character())
p1
```
```{r echo = FALSE}
5457
```

--- Part 2 ---

```{r eval = FALSE}
connections <- input

paths <- connections |> 
  bind_rows(connections |> 
              select(left = right, right = left))

n <- "start"
p <- ""

valid_child <- function(n,p){
  # no start!
  if(n == "start") return(FALSE)
  # if haven't been there, is ok
  if(!n %in% p) return(TRUE)
  # if is uppercase, ok to go back
  if(identical(toupper(n),n)) return(TRUE)
  # if not uppercase and have been there only once, ok 
  lower_p <- p[tolower(p)==p]
  if(identical(lower_p,unique(lower_p))) return(TRUE)
  # otherwise bad.
  return(FALSE)
}

pathfinder <- function(n, current_path){
  current_path <- c(current_path,n)
  
  if(n == "end") return(tibble(path = list(current_path)))
  
  child_nodes <- paths$right[paths$left == n]
  
  valid_child_nodes <- child_nodes[map_lgl(child_nodes,
                                           valid_child, 
                                           current_path)]
  
  if(length(valid_child_nodes) == 0) return(NULL)
  
  all_paths <- tibble(path = list())
  
  for(cn in valid_child_nodes) {
    all_paths <- bind_rows(all_paths,pathfinder(cn,current_path))
  }
  
  return(all_paths)
}

p2 <- pathfinder(n = "start", current_path = character())
nrow(p2)
```
```{r echo = FALSE}
128506
```

