---
title: 'Advent Of Code: 2021-10'
author: "Tan Ho"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(glue)
  
  knitr::opts_chunk$set(echo = TRUE)
})

options(scipen = 9999999)
options(dplyr.summarise.inform = FALSE)
```

--- Data ---

```{r eval = FALSE}
# tanho63/aoc.elf
aoc.elf::aoc_get(day = 10, year = 2021)
```

```{r}
input <- readLines(here::here("2021/day-10-input.txt")) |> 
  strsplit("")

```

--- Part 1 ---

In a loop, collected a queue of open-brackets and then check each closing bracket against the last item in the opening queue. If check fails, exit early with the failure. 

```{r}
checker <- function(v){
  
  v <- unlist(v)
  
  open_char <- c("{","[","(","<")
  close_char <- c("}", "]", ")", ">")
  
  mapping <- tibble(open = open_char,
                    close = close_char) |> 
    deframe()
  
  queue <- c()
  
  for(i in v){
    if(i %in% open_char) queue <- c(queue,i)
    if(i %in% close_char) {
      check_closing <- mapping[tail(queue,1)] == i
      if(!check_closing) return(i)
      queue <- head(queue,-1)
    }
  }
  
  return("OK")
}

x <- map(input,checker) |> unlist()

value_map <- tibble(
  close_char = c("}", "]", ")", ">"),
  value = c(1197,57, 3, 25137)
) |> 
  deframe()

sum(value_map[x[x != "OK"]])
```

--- Part 2 ---

Did myself a massive favour here with my part one algo - already had a queue of opening brackets and the code to fetch the closing brackets, so just had to return the closing brackets for the queue.

```{r}

p2 <- input[unlist(x)== "OK"]

completer <- function(v){
  
  v <- unlist(v)
  
  open_char <- c("{","[","(","<")
  close_char <- c("}", "]", ")", ">")
  
  mapping <- tibble(open = open_char,
                    close = close_char) |> 
    deframe()
  
  queue <- c()
  
  for(i in v){
    if(i %in% open_char) queue <- c(queue,i)
    if(i %in% close_char) {
      check_closing <- mapping[tail(queue,1)] == i
      if(!check_closing) return(i)
      queue <- head(queue,-1)
    }
  }
  
  return(mapping[rev(queue)])
  
}

x <- map(p2,completer)

scorer <- function(x){
  score <- 0
  value_table <- tibble(
    char = c("}", "]", ")", ">"),
    score = c(3,2,1,4)
  ) |> 
    deframe()
  
  for(i in x){
    
    score <- score * 5
    score <- score + value_table[i]
  }
  
  return(score)
}


map(x,scorer) |> unlist() |> median()

```

