---
title: 'Advent Of Code: 2021-15'
author: Tan Ho
date: "`r Sys.Date()`"
output: html_document
---

```{r eval = FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(glue)
  
  knitr::opts_chunk$set(echo = TRUE)
})

options(scipen = 9999999)
options(dplyr.summarise.inform = FALSE)
```

--- Data ---

```{r eval = FALSE}
# tanho63/aoc.elf
aoc.elf::aoc_get(day = 15, year = 2021)
```

```{r eval = FALSE}
input <- readLines(here::here("2021/day-15-input.txt")) |> 
  strsplit("") |> 
  map(as.numeric) |> 
  reduce(rbind)

example <- c("1163751742",
           "1381373672",
           "2136511328",
           "3694931569",
           "7463417111",
           "1319128137",
           "1359912421",
           "3125421639",
           "1293138521",
           "2311944581") |> 
  strsplit("") |> 
  map(as.numeric) |> 
  reduce(rbind)

```

--- Part 1 ---

```{r eval = FALSE}
# give me all available up/down/left/right steps that are on the grid

m <- input

get_neighbours <- function(row_loc, col_loc, queue){
  
  n <- tibble(row = c(-1,1,0,0), col = c(0,0,1,-1)) |> 
    mutate(
      row = row + row_loc,
      col = col + col_loc
    ) |> 
    filter(row <= nrow(m), 
           col <= nrow(m), 
           row >= 1, 
           col >= 1) |> 
    inner_join(queue, by = c("row","col"))
  
  return(n)
}
  
  # populate distance matrix of equal size to m
  distances <- matrix(Inf,nrow = nrow(m),ncol = ncol(m))
  
  # set origin as top left hand corner
  distances[1] <- 0
  
  # create queue of indices to check
  queue <- crossing(row = 1:nrow(m), col = 1:ncol(m))
  
  # create copy of queue that will list the shortest path from source.
  prev <- queue |> 
    mutate(prev = 0)
  
  counter <- 0
  
  while(nrow(queue) > 0){
    
    counter <- counter + 1
    
    queue <- queue |> 
      mutate(dist = distances[cbind(row,col)])
    
    i <- queue |> slice_min(dist,with_ties = FALSE,n = 1)
    
    queue <- queue |> anti_join(i, by = c("row","col"))
    
    neighbours <- get_neighbours(i$row, i$col, queue)
    
    if(nrow(neighbours) == 0) next
    
    for(n in seq_len(nrow(neighbours))){
      
      possible_dist <- i$dist + m[neighbours$row[n],neighbours$col[n]]
      
      if(possible_dist < distances[neighbours$row[n],neighbours$col[n]]){
        distances[neighbours$row[n],neighbours$col[n]] <- possible_dist
        prev$prev[prev$row == neighbours$row[n] & prev$col == neighbours$col[n]] <- counter
      }
      
    }
    
  }

  distances[nrow(m),ncol(m)]
  
```
```{r echo = FALSE}
523
```

--- Part 2 ---

Ran into a bunch of problems and ended up leaving on overnight after attempting some optimizations.

Build out incremented matrix.

```{r eval = FALSE}
m <- input

col_1 <- map(0:4,~ifelse(m + .x > 9,m + .x - 9, m + .x)) |> reduce(rbind)
col_2 <- map(1:5,~ifelse(m + .x > 9,m + .x - 9, m + .x)) |> reduce(rbind)
col_3 <- map(2:6,~ifelse(m + .x > 9,m + .x - 9, m + .x)) |> reduce(rbind)
col_4 <- map(3:7,~ifelse(m + .x > 9,m + .x - 9, m + .x)) |> reduce(rbind)
col_5 <- map(4:8,~ifelse(m + .x > 9,m + .x - 9, m + .x)) |> reduce(rbind)

m <- cbind(col_1,col_2,col_3,col_4,col_5)
```

build out a list structure of neighbours rather than calculating at each step

```{r eval = FALSE}
neighbours_matrix <- crossing(row = seq_len(nrow(m)),
                              col = seq_len(ncol(m))) |> 
  arrange(col,row) |> 
  mutate(index = row_number(),
         add = list(tibble(rp = c(-1,1,0,0), cp = c(0,0,1,-1)))
  ) |> 
  unnest(add) |> 
  mutate(row2 = row + rp,
         col2 = col + cp) |> 
  filter(row2 <= nrow(m),
         col2 <= ncol(m),
         row2 >=1,
         col2 >=1) |> 
  select(-rp,-cp) %>%
  left_join(.,.,by = c("row2"="row","col2"="col")) |> 
  select(index = index.x, neighbour = index.y) |> 
  distinct() |> 
  group_by(index) |> 
  summarise(neighbour = list(neighbour)) |> 
  deframe()

head(neighbours_matrix)
```
```{r echo = FALSE}
list(`1` = c(2L, 501L), `2` = c(1L, 3L, 502L), `3` = c(2L, 4L, 
503L), `4` = c(3L, 5L, 504L), `5` = c(4L, 6L, 505L), `6` = c(5L, 
7L, 506L))
```

run dijkstra's algo (?) oddly slow.

```{r eval = FALSE}
  distances <- matrix(Inf,nrow = nrow(m),ncol = ncol(m))
  
  # set origin as top left hand corner
  distances[1] <- 0
  
  # create queue of indices to check
  queue <- seq_along(m) |> as.list()
  
  counter <- 0
  
  while(length(queue) > 0){
    
    counter <- counter + 1
    
    if(counter %% 1000 == 0) cat(counter, format(Sys.time()), "\n")
    
    i <- queue[which.min(distances[queue])] # slow?
    
    queue <- queue[queue!=i] #slow?
    
    neighbours <- neighbours_matrix[[as.character(i)]]
    neighbours <- neighbours[neighbours %in% queue] # slow?
    
    for(n in neighbours){
      
      possible_dist <- distances[i] + m[n] 
      
      if(possible_dist < distances[n]) distances[n] <- possible_dist
    }
  }

  distances[nrow(m),ncol(m)]
```
```{r echo = FALSE}
2876
```

--- Optimizing ---

Optimizing. Biggest culprits so far: 

- `neighbours %in% queue` scans 250K vector for ~1-4 objects, much better is a visited matrix to lookup by position. net gain: 45 minutes -> 6 minutes (screaming internally.)
- `queue <- queue != i` should be next? 

```{r eval = FALSE}
# populate distance matrix of equal size to m
  distances <- matrix(Inf,nrow = nrow(m),ncol = ncol(m))
  
  # set origin as top left hand corner
  distances[1] <- 0
  
  # create queue of indices to check
  queue <- seq_along(m)
  visited <- matrix(F,nrow = nrow(m), ncol = ncol(m))
  counter <- 0
  
  while(!visited[nrow(m),ncol(m)]){
    
    counter <- counter + 1
    
    if(counter %% 1000 == 0) cat(counter, format(Sys.time()), "\n")
    
    i <- queue[which.min(distances[queue])] # slow?
    queue <- queue[queue!=i] #slow?
    visited[i] <- TRUE
    
    neighbours <- neighbours_matrix[i] |> unlist()
    neighbours <- neighbours[!visited[neighbours]]
    
    for(n in neighbours){
      
      possible_dist <- distances[i] + m[n]
      if(possible_dist < distances[n]) distances[n] <- possible_dist
    }
  }

  distances[nrow(m),ncol(m)]
```
```{r echo = FALSE}
2876
```
